---
title: "Step by step guide for creating a package that depends on RStan"
author: "Stefan Siegert, Jonah Gabry, Martin Lysy, and Ben Goodrich"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Step by step guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r SETTINGS-knitr, include=FALSE}
stopifnot(require(knitr))
opts_chunk$set(
  comment=NA,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
td <- tempdir()
PATH <- file.path(td, "rstanlm")
if(dir.exists(PATH)) {
  unlink(PATH, recursive = TRUE, force = TRUE)
}
```

## Introduction

In this vignette we will walk through the steps necessary for creating an
R package that depends on Stan by creating a package with one function
that fits a simple linear regression. Before continuing, we recommend that you first read the other vignette [_Guidelines for Developers of R Packages Interfacing with Stan_](https://mc-stan.org/rstantools/articles/developer-guidelines.html).

## Creating the package skeleton

The **rstantools** package offers two methods for adding Stan functionality
to R packages:

  * `rstan_create_package()`: set up a new R package with Stan programs
  * `use_rstan()`: add Stan functionality to an _existing_ R package

Here we will use `rstan_create_package()` to initialize a bare-bones package
directory. The name of our demo package will be __rstanlm__; it will fit a
simple linear regression model using Stan.

```{r rstan_create_package, eval=FALSE}
library("rstantools")
rstan_create_package(path = 'rstanlm')
```
```{r rstan_create_package-eval, echo=FALSE,warning=FALSE}
library("rstantools")
rstan_create_package(path = PATH, rstudio=FALSE, open=FALSE)
```

If we had existing `.stan` files to include with the package we could use the
optional `stan_files` argument to `rstan_create_package()` to include them.
Another option, which we'll use below, is to add the Stan files once the
basic structure of the package is in place.

We can now set the new working directory to the new package directory and view
the contents. (Note: if using RStudio then by default the newly created project
for the package will be opened in a new session and you will not need the call
to `setwd()`.)


```{r, eval=FALSE}
setwd("rstanlm")
list.files(all.files = TRUE)
```
```{r, echo=FALSE}
list.files(PATH, all.files = TRUE)
```

```{r, eval=FALSE}
file.show("DESCRIPTION")
```
```{r, echo=FALSE}
DES <- readLines(file.path(PATH, "DESCRIPTION"))
cat(DES, sep = "\n")
```

Some of the sections in the `DESCRIPTION` file need to be edited by hand (e.g.,
`Title`, `Author`, `Maintainer`, and `Description`, but these also can be set
with the `fields` argument to `rstan_create_package()`).  However,
`rstan_create_package()` has added the necessary packages and versions to
`Depends`, `Imports`, and `LinkingTo` to enable Stan functionality.

## Read-and-delete-me file

Before deleting the `Read-and-delete-me` file in the new package directory make
sure to read it because it contains some important instructions about
customizing your package:

```{r, eval=FALSE}
file.show("Read-and-delete-me")
```
```{r, echo=FALSE}
cat(readLines(file.path(PATH, "Read-and-delete-me")), sep = "\n")
```

You can move this file out of the directory, delete it, or list it in the
`.Rbuildignore` file if you want to keep it in the directory.

```{r, eval=FALSE}
file.remove('Read-and-delete-me')
```
```{r, echo=FALSE}
file.remove(file.path(PATH, 'Read-and-delete-me'))
```


## Stan files

Our package will call **rstan**'s `sampling()` method to use MCMC to fit a simple
linear regression model for an outcome variable `y` with a single predictor `x`.
After writing the necessary Stan program, the file should be saved with a
`.stan` extension in the `inst/stan` subdirectory. We'll save the
following program to `inst/stan/lm.stan`:

```{stan, output.var = "foo", eval = FALSE}
// Save this file as inst/stan/lm.stan
data {
  int<lower=1> N;
  vector[N] x;
  vector[N] y;
}
parameters {
  real intercept;
  real beta;
  real<lower=0> sigma;
}
model {
  // ... priors, etc.

  y ~ normal(intercept + beta * x, sigma);
}

```

```{r, include=FALSE}
stan_prog <- "
data {
  int<lower=1> N;
  vector[N] x;
  vector[N] y;
}
parameters {
  real intercept;
  real beta;
  real<lower=0> sigma;
}
model {
  // ... priors, etc.

  y ~ normal(intercept + beta * x, sigma);
}
"
writeLines(stan_prog, con = file.path(PATH, "inst", "stan", "lm.stan"))
rstan_config(PATH)
```

The `inst/stan` subdirectory can contain additional Stan programs if
required by your package. During installation, all Stan programs will be
compiled and saved in the list `stanmodels` that can then be used by R function
in the package. The rule is that the Stan program compiled from the model code
in `inst/stan/foo.stan` is stored as list element `stanmodels$foo`. Thus, the
filename of the Stan program in the `inst/stan` directory should not contain
spaces or dashes and nor should it start with a number or utilize non-ASCII
characters.

## R files

We next create the file `R/lm_stan.R` where we define the function `lm_stan()`
in which our compiled Stan model is being used.  Setting the
`rstan_create_package()` argument `roxygen = TRUE` (the default value) enables
[__roxygen2__](https://CRAN.R-project.org/package=roxygen2) documentation for
the package functions.  The following comment block placed in `lm_stan.R`
ensures that the function has a help file and that it is added to the package
`NAMESPACE`:

```{r}
# Save this file as `R/lm_stan.R`

#' Bayesian linear regression with Stan
#'
#' @export
#' @param x Numeric vector of input values.
#' @param y Numeric vector of output values.
#' @param ... Arguments passed to `rstan::sampling` (e.g. iter, chains).
#' @return An object of class `stanfit` returned by `rstan::sampling`
#'
lm_stan <- function(x, y, ...) {
  standata <- list(x = x, y = y, N = length(y))
  out <- rstan::sampling(stanmodels$lm, data = standata, ...)
  return(out)
}

```


```{r, include=FALSE}
Rcode <- "
#' Bayesian linear regression with Stan
#'
#' @export
#' @param x Numeric vector of input values.
#' @param y Numeric vector of output values.
#' @param ... Arguments passed to `rstan::sampling`.
#' @return An object of class `stanfit` returned by `rstan::sampling`
lm_stan <- function(x, y, ...) {
  out <- rstan::sampling(stanmodels$lm, data=list(x=x, y=y, N=length(y)), ...)
  return(out)
}
"
writeLines(Rcode, con = file.path(PATH, "R", "lm_stan.R"))
```

When __roxygen2__ documentation is enabled, a top-level package file
`R/rstanlm-package.R` is created by `rstan_create_package()` to import necessary
functions for other packages and to set up the package for compiling Stan C++
code:

```{r, eval=FALSE}
file.show(file.path("R", "rstanlm-package.R"))
```
```{r, echo=FALSE}
cat(readLines(file.path(PATH, "R", "rstanlm-package.R")), sep = "\n")
```

The `#' @description` section can be manually edited to provided specific
information about the package.

## Documentation

With __roxygen__ documentation enabled, we need to generate the documentation
for `lm_stan` and update the `NAMESPACE` so the function is exported, i.e.,
available to users when the package is installed.  This can be done with the
function `roxygen2::roxygenize()`, which needs to be called twice initially.

```{r, eval = FALSE}
try(roxygen2::roxygenize(load_code = rstantools_load_code), silent = TRUE)
roxygen2::roxygenize()
```
```{r, echo=FALSE, results="hide"}
try(roxygen2::roxygenize(PATH, load_code = rstantools_load_code), silent = TRUE)
roxygen2::roxygenize(PATH)
```

## Install and use

Finally, the package is ready to be installed:

```{r,eval=FALSE}
# using ../rstanlm because already inside the rstanlm directory
install.packages("../rstanlm", repos = NULL, type = "source")
```
```{r,echo=FALSE}
install.packages(PATH, repos = NULL, type = "source")
```

It is also possible to use `devtools::install(quick=FALSE)` to install the
package. The argument `quick=FALSE` is necessary if you want to recompile the
Stan models. Going forward, if you only make a  change to the R code or the
documentation, you can set `quick=TRUE` to speed up the process, or use
`devtools::load_all()`.

After installation, the package can be loaded and used like any other R package:

```{r, eval=FALSE}
library("rstanlm")
```

```{r}
fit <- lm_stan(y = rnorm(10), x = rnorm(10), 
               # arguments passed to sampling
               iter = 2000, refresh = 500)
print(fit)
```

```{r, echo=FALSE}
unlink(PATH, recursive = TRUE, force = TRUE)
```

## Advanced options

Details can be found in the documentation for `rstan_create_package()` so we
only mention some of these briefly here:

* Running `rstan_create_package()` with `auto_config = TRUE` (the default value)
automatically synchronizes the Stan C++ files with the `.stan` model files
located in `inst/stan`, although this creates a dependency of your package on
__rstantools__ itself (i.e., __rstantools__ must be installed for your package
to work).  Setting `auto_config = FALSE` removes this dependency, at the cost of
having to manually synchronize Stan C++ files by running `rstan_config()` every
time a package `.stan` file is added, removed, or even just modified.

* The function `use_rstan()` can be used to add Stan functionality to an
existing package, instead of building the package from scratch.
  * Note: If you are already using roxygen in your package, you'll have to use
    roxygen to update your Namespace file via the `R/<package-name>-package.R` file.
    Check the roxygen documentation for more details.

## Adding additional Stan models to an existing R package with Stan models

One may add additional Stan models to an existing package.
The following steps are required if one is using `devtools`:

1.  Add new Stan file, e.g., `inst/stan/new.stan`
2.  Run `pkgbuild::compile_dll()` to preform a fake R CMD install.
3.  Run  `roxygen2::roxygenize()` to update the documentation.
4.  Run ` devtools::install()` to install the package locally.

## Links

* [_Guidelines for Developers of R Packages Interfacing with Stan_](https://mc-stan.org/rstantools/articles/developer-guidelines.html)

* Ask a question at the [Stan Forums](https://discourse.mc-stan.org/)

* [R packages](https://r-pkgs.org/) by Hadley Wickham and Jenny Bryan provides
a solid foundation in R package development as well as the release process.

